Gaze-dependent evidence accumulation predicts multi-alternative risky choice behaviour

### 2021-02-13 manuscript/manuscript_refs-apa.tex
minor fixes

While current gaze-dependent accumulation models of risky choice assume that each alternative can be assigned a singular underlying scale value (e.g., expected utilities for each alternative), choices are also assumed to be constructed in an accumulation process that remains malleable due to its dependence on gaze. Therefore, gaze-dependent accumulation models potentially could account for violations of IIA or regularity if the distribution of gaze itself changed depending on the decision context (e.g., shift towards dominant alternatives in the attraction effect, compromise alternatives in the attraction effect, or dissimilar alternatives in the similarity effect). Initial evidence for such an attention-shift in the attraction effect comes from a study of intertemporal choice \parencite{marini2020AttractionComesMany}. Furthermore, recent work has suggested gaze-dependent evidence accumulation as a model of multi-alternative, multi-attribute choice in the absence of context effects \parencite{cohen2017MultiattributeMultialternativeModels,gluth2018ValuebasedAttentionalCapture}.

@@

In our experiment, 40 participants made repeated decisions between three all-or-nothing gambles, each described by a probability $p$ to win an outcome $m$ and nothing otherwise (Fig. \ref{fig:task}a). We recorded participants' eye movements during the task with an eye tracker (see \nameref{sec:methods} for details). Participants were instructed that after completing the task, their chosen gamble from one randomly determined trial would be played out for an additional bonus payment. In contrast to hypothetical choices between consumer goods described on attribute dimensions like quality and price, risky gamble stimuli offer a high amount of control over their attributes and a straightforward way to incentivise choices. The gamble stimuli were designed to elicit attraction and compromise effects and were individually tailored to account for each participant's risk preferences (Fig. \ref{fig:task}b). Participants performed three pairs of indifference-estimation and experimental blocks, for a total of 225 experimental trials (see \nameref{sec:methods} for additional details on the experimental procedure and gamble stimuli). In indifference-estimation blocks, participants made repeated choices between pairs of gambles with different probabilities (Fig. \ref{fig:task}c). Gamble outcomes $m$ were adjusted according to participants' choices so that four approximately equally preferred gambles $C_B$, $B$, $A$, and $C_A$ were constructed with winning probabilities $p$ = 55\%, 65\%, 75\%, and 85\%. Additionally, two asymmetrically dominated decoys $D_A$ and $D_B$ were defined to be 2\% and 1 EUR worse than gambles $A$ and $B$, respectively. Following this indifference estimation, participants performed an experimental block of 75 ternary choice trials, 32 of which were compromise trials balanced with respect to the target option (i.e., 16 choice sets \{$A$, $B$, $C_A$\}, 16 choice sets \{$A$, $B$, $C_B$\}), 32 attraction trials (16 choice sets \{$A$, $B$ ,$D_A$\}, 16 choice sets \{$A$, $B$, $D_B$\}) and 11 distractor trials showing randomly created options with expected value of 10 EUR and low (5–33\%), medium (34–64\%) and high (65–95\%) probability $p$. Uniform noise was added to all outcomes $m$ ($\pm$0 EUR, +1 EUR) and probabilities $p$ (–3\%, $\pm$0\%, +3\%) in each trial. Asymmetrically dominated decoys $D_A$ and $D_B$ received the same noise as the target option to preserve the dominance relation.

@@

\subsection*{Behavioural modeling and model comparison}

@@

All gaze-dependent models were able to capture this positive association. Note, however, that they also predicted choices of dominated decoys in attraction trials, if decoys had a large dwell time advantage (Fig. 3g). In this case, GLA performed better than GB$_{\text{dyn}}$ and GB$_{\text{stat}}$, as it predicted fewer decoy choices. However, MDFT and EU generally could not capture the empirical association of gaze and choice; they predicted too many choices of alternatives with negative dwell time advantage, and too few choices of alternatives with positive dwell time advantage. 
Nor did they predict choices of dominated decoys, even if the decoy was looked at longer.

@@

On the level of model mechanisms, multiplicative attribute integration outperformed additive integration (mean BIC$_{\text{mul.}}$ = 312.25, mean BIC$_{\text{add.}}$ = 333.20; Fig. 4b), inclusion of an alternative-wise gaze-discount (mean BIC$_{\text{GD alt.}}$ = 289.13, mean BIC$_{\text{no GD alt.}}$  = 356.32), but not attribute-wise gaze-discount yielded lower BIC scores (mean BIC$_{\text{GD att.}}$  = 324.11, mean BIC$_{\text{no GD att.}}$ = 321.33). Other gaze-dependent mechanisms also improved model fit: Variants with gaze-dependent leak yielded lower BIC scores than variants with constant or no leak (BIC$_{\text{gaze}}$  = 292.06, BIC$_{\text{constant}}$ = 304.84, BIC$_{\text{none}}$ = 381.49). Gaze-dependent inhibition performed better than constant, none or distance-dependent inhibition (BIC$_{\text{gaze}}$ = 305.64, BIC$_{\text{constant}}$ = 316.06, BIC$_{\text{none}}$ = 311.75, BIC$_{\text{distance}}$ = 351.57). In summary, all mechanisms that allowed a model to exploit the positive association between gaze and choice improved model fit on average.

@@

Finally, we analysed the hybrid model variant in more detail (variant B in Table \ref{tab:switchboard-best-ind}), which described 9 participants best. This variant performed especially well for participants with large attraction effects (Fig. \ref{fig:hybrid}a), whereas GLA best described the majority of participants with attraction RST around 0.5. In contrast, better-performing variants could not be separated by compromise effect strength (Fig. \ref{fig:hybrid}b). The hybrid variant correctly predicted individual differences in the attraction effect (correlation between observed and predicted RST: $r$ = 0.92, HDI$_{95}$ = [0.86 0.96]; Fig. \ref{fig:hybrid}c). Here, it performed better than the GLA model (Fig. \ref{fig:modelcomparison}i), as it was also able to capture the behaviour of participants with particularly strong attraction effects: Using distance-dependent inhibition, it was able to predict high choice probabilities for target alternatives in attraction trials for some participants, and fewer choices of competitor and dominated decoy alternatives (Supplementary Fig. 4f). Predictions of individual RST in compromise trials were almost identical between the two models ($r$ = 0.81, HDI$_{95}$ = [0.70, 0.91]; Fig. \ref{fig:hybrid}d, see Fig. \ref{fig:modelcomparison}j for GLA).

@@

where $k_m$ is the number of free parameters of model $m$ and $\ln{\hat{L}}$ is the summed log-likelihood of observed choices under model $m$.

@@

\subsubsection*{Attribute-wise gaze discount}
The analysis also included the option of attribute-wise gaze-dependent discounting (similar to the two-layer model from \parencite{glickman2019FormationPreferenceRisky} and the model presented in \parencite{fisher2017AttentionalDriftDiffusion}. If switched on, stimulus attributes of the currently unattended dimension (e.g., probability, when an lottery outcome was fixated) were discounted by a free parameter $\eta$ ($0 \le \eta \le 1$). In combination with \emph{additive} attribute integration, the attribute-wise gaze discount was applied after attribute normalization, but prior to the weighted addition. For \emph{multiplicative} attribute integration, attributes were discounted before entering probability-weighting and utility functions $w$ and $U$.

@@

Regression weight estimates are shown in Supplementary Fig. \ref{fig:dwell-regression-weights}. Across the trial, we find strong effects of position on dwell time, so that dwell times towards the top and left were longer. Significant negative interaction effects of the row and column predictors with time showed that these effects diminish across the trial. We also find a gaze-cascade effect \parencite{mullett2016ImplicationsVisualAttention,shimojo2003GazeBiasBoth}, where dwell times to AOIs belonging to the ultimately chosen alternative are longer across the trial (and increasingly so throughout the trial, indicated by the positive interaction term with time). Dwell times to decoys decreased significantly during the trial, and across the full trial, dwell times to decoys were shorter than other alternatives. Similarly, dwell times towards probability attributes p shortened across the trial. Across the full trial, however, dwell times towards probability AOIs were not shorter than those to outcome AOIs. Finally, dwell times to target alternatives were longer across the trial in both compromise and attraction trials. In addition, this effect increased throughout the trial in compromise, but not in attraction trials. Note that these effects are independent of the effect of choice, as choice is a separate predictor in the model. We could not find an association between the attribute rank (being the worst, best, or intermediate item on an attribute) and dwell time.

>>>>

While current gaze-dependent accumulation models of risky choice assume that each alternative can be assigned a singular underlying scale value (e.g., expected utilities for each alternative), choices are also assumed to be constructed in an accumulation process that remains malleable due to its dependence on gaze. Therefore, gaze-dependent accumulation models potentially could account for violations of IIA if the distribution of gaze itself changed depending on the decision context (e.g., shift towards dominant alternatives in the attraction effect, compromise alternatives in the attraction effect, or dissimilar alternatives in the similarity effect). Initial evidence for such an attention-shift in the attraction effect comes from a study of intertemporal choice \parencite{marini2020AttractionComesMany}. Furthermore, recent work has suggested gaze-dependent evidence accumulation as a model of multi-alternative, multi-attribute choice in the absence of context effects \parencite{cohen2017MultiattributeMultialternativeModels,gluth2018ValuebasedAttentionalCapture}.

@@

In our experiment, 40 participants made repeated decisions between three all-or-nothing gambles, each described by a probability $p$ to win an outcome $m$ and nothing otherwise (Fig. \ref{fig:task}a). We recorded participants' eye movements during the task with an eye tracker (see \nameref{sec:methods} for details). Participants were instructed that after completing the task, their chosen gamble from one randomly determined trial would be played out for an additional bonus payment. In contrast to hypothetical choices between consumer goods described on attribute dimensions like quality and price, risky gamble stimuli offer a high amount of control over their attributes and a straightforward way to incentivise choices. The gamble stimuli were designed to elicit attraction and compromise effects and were individually tailored to account for each participant's risk preferences (Fig. \ref{fig:task}b). Participants performed three pairs of indifference-estimation and experimental blocks, for a total of 225 experimental trials (see \nameref{sec:methods} for additional details on the experimental procedure and gamble stimuli). In indifference-estimation blocks, participants made repeated choices between pairs of gambles with different probabilities (Fig. \ref{fig:task}c). Gamble outcomes $m$ were adjusted according to participants' choices so that four approximately equally preferred gambles $C_B$, $B$, $A$, and $C_A$ were constructed with winning probabilities $p$ = 55\%, 65\%, 75\%, and 85\%. Additionally, two asymmetrically dominated decoys $D_A$ and $D_B$ were defined to be 2\% and 1 EUR worse than gambles $A$ and $B$, respectively. Following this indifference estimation, participants performed an experimental block of 75 ternary choice trials, 32 of which were compromise trials balanced with respect to the target option (i.e., 16 choice sets \{$A$, $B$, $C_A$\}, 16 choice sets \{$A$, $B$, $C_B$\}), 32 attraction trials (16 choice sets \{$A$, $B$, $D_A$\}, 16 choice sets \{$A$, $B$, $D_B$\}) and 11 distractor trials showing randomly created options with expected value of 10 EUR and low (5–33\%), medium (34–64\%) and high (65–95\%) probability $p$. Uniform noise was added to all outcomes $m$ ($\pm$0 EUR, +1 EUR) and probabilities $p$ (–3\%, $\pm$0\%, +3\%) in each trial. Asymmetrically dominated decoys $D_A$ and $D_B$ received the same noise as the target option to preserve the dominance relation.

@@

\subsection*{Behavioural modelling and model comparison}

@@

All gaze-dependent models were able to capture this positive association. Note, however, that they also predicted choices of dominated decoys in attraction trials, if decoys had a large dwell time advantage (Fig. \ref{fig:modelcomparison}g). In this case, GLA performed better than GB$_{\text{dyn}}$ and GB$_{\text{stat}}$, as it predicted fewer decoy choices. However, MDFT and EU generally could not capture the empirical association of gaze and choice; they predicted too many choices of alternatives with negative dwell time advantage, and too few choices of alternatives with positive dwell time advantage. 
Nor did they predict choices of dominated decoys, even if the decoy was looked at longer.

@@

On the level of model mechanisms, multiplicative attribute integration outperformed additive integration (mean BIC$_{\text{mul.}}$ = 312.25, mean BIC$_{\text{add.}}$ = 333.20; Fig. \ref{fig:switchboard}b), inclusion of an alternative-wise gaze-discount (mean BIC$_{\text{GD alt.}}$ = 289.13, mean BIC$_{\text{no GD alt.}}$  = 356.32), but not attribute-wise gaze-discount yielded lower BIC scores (mean BIC$_{\text{GD att.}}$  = 324.11, mean BIC$_{\text{no GD att.}}$ = 321.33). Other gaze-dependent mechanisms also improved model fit: Variants with gaze-dependent leak yielded lower BIC scores than variants with constant or no leak (BIC$_{\text{gaze}}$  = 292.06, BIC$_{\text{constant}}$ = 304.84, BIC$_{\text{none}}$ = 381.49). Gaze-dependent inhibition performed better than constant, none or distance-dependent inhibition (BIC$_{\text{gaze}}$ = 305.64, BIC$_{\text{constant}}$ = 316.06, BIC$_{\text{none}}$ = 311.75, BIC$_{\text{distance}}$ = 351.57). In summary, all mechanisms that allowed a model to exploit the positive association between gaze and choice improved model fit on average.

@@

Finally, we analysed the hybrid model variant in more detail (variant B in Table \ref{tab:switchboard-best-ind}), which described 9 participants best. This variant performed especially well for participants with large attraction effects (Fig. \ref{fig:hybrid}a), whereas GLA best described the majority of participants with attraction RST around 0.5. In contrast, better-performing variants could not be separated by compromise effect strength (Fig. \ref{fig:hybrid}b). The hybrid variant correctly predicted individual differences in the attraction effect (correlation between observed and predicted RST: $r$ = 0.92, HDI$_{95}$ = [0.86 0.96]; Fig. \ref{fig:hybrid}c). Here, it performed better than the GLA model (Fig. \ref{fig:modelcomparison}i), as it was also able to capture the behaviour of participants with particularly strong attraction effects: Using distance-dependent inhibition, it was able to predict high choice probabilities for target alternatives in attraction trials for some participants, and fewer choices of competitor and dominated decoy alternatives (Supplementary Fig. \ref{fig:model-pred-probs}f). Predictions of individual RST in compromise trials were almost identical between the two models ($r$ = 0.81, HDI$_{95}$ = [0.70, 0.91]; Fig. \ref{fig:hybrid}d, see Fig. \ref{fig:modelcomparison}j for GLA).

@@

where $k_m$ is the number of free parameters of model $m$ and $\ln{(\hat{L})}$ is the summed log-likelihood of observed choices under model $m$.

@@

\subsubsection*{Attribute-wise gaze discount}
The analysis also included the option of attribute-wise gaze-dependent discounting (similar to the two-layer model from \textcite{glickman2019FormationPreferenceRisky} and the model presented in \textcite{fisher2017AttentionalDriftDiffusion}). If switched on, stimulus attributes of the currently unattended dimension (e.g., probability, when an lottery outcome was fixated) were discounted by a free parameter $\eta$ ($0 \le \eta \le 1$). In combination with \emph{additive} attribute integration, the attribute-wise gaze discount was applied after attribute normalization, but prior to the weighted addition. For \emph{multiplicative} attribute integration, attributes were discounted before entering probability-weighting and utility functions $w$ and $U$.

@@

Regression weight estimates are shown in Supplementary Fig. \ref{fig:dwell-regression-weights}. Across the trial, we find strong effects of position on dwell time, so that dwell times towards the top and left were longer. Significant negative interaction effects of the row and column predictors with time showed that these effects diminish across the trial. We also find a gaze-cascade effect \parencite{mullett2016ImplicationsVisualAttention,shimojo2003GazeBiasBoth}, where dwell times to AOIs belonging to the ultimately chosen alternative are longer across the trial (and increasingly so throughout the trial, indicated by the positive interaction term with time). Dwell times to decoys decreased significantly during the trial, and across the full trial, dwell times to decoys were shorter than other alternatives. Similarly, dwell times towards probability attributes $p$ shortened across the trial. Across the full trial, however, dwell times towards probability AOIs were not shorter than those to outcome AOIs. Finally, dwell times to target alternatives were longer across the trial in both compromise and attraction trials. In addition, this effect increased throughout the trial in compromise, but not in attraction trials. Note that these effects are independent of the effect of choice, as choice is a separate predictor in the model. We could not find an association between the attribute rank (being the worst, best, or intermediate item on an attribute) and dwell time.

### 2021-02-13 manuscript/manuscript_refs-numbered.tex
minor fixes

While current gaze-dependent accumulation models of risky choice assume that each alternative can be assigned a singular underlying scale value (e.g., expected utilities for each alternative), choices are also assumed to be constructed in an accumulation process that remains malleable due to its dependence on gaze. Therefore, gaze-dependent accumulation models potentially could account for violations of IIA or regularity if the distribution of gaze itself changed depending on the decision context (e.g., shift towards dominant alternatives in the attraction effect, compromise alternatives in the attraction effect, or dissimilar alternatives in the similarity effect). Initial evidence for such an attention-shift in the attraction effect comes from a study of intertemporal choice \autocite{marini2020AttractionComesMany}. Furthermore, recent work has suggested gaze-dependent evidence accumulation as a model of multi-alternative, multi-attribute choice in the absence of context effects \autocite{cohen2017MultiattributeMultialternativeModels,gluth2018ValuebasedAttentionalCapture}.

@@

\subsubsection*{Attribute-wise gaze discount}
The analysis also included the option of attribute-wise gaze-dependent discounting (similar to the two-layer model from \citeauthor{glickman2019FormationPreferenceRisky} \autocite{glickman2019FormationPreferenceRisky} and the model presented in \citeauthor{fisher2017AttentionalDriftDiffusion} \autocite{fisher2017AttentionalDriftDiffusion}. If switched on, stimulus attributes of the currently unattended dimension (e.g., probability, when an lottery outcome was fixated) were discounted by a free parameter $\eta$ ($0 \le \eta \le 1$). In combination with \emph{additive} attribute integration, the attribute-wise gaze discount was applied after attribute normalization, but prior to the weighted addition. For \emph{multiplicative} attribute integration, attributes were discounted before entering probability-weighting and utility functions $w$ and $U$.

@@

Regression weight estimates are shown in Supplementary Fig. \ref{fig:dwell-regression-weights}. Across the trial, we find strong effects of position on dwell time, so that dwell times towards the top and left were longer. Significant negative interaction effects of the row and column predictors with time showed that these effects diminish across the trial. We also find a gaze-cascade effect \autocite{mullett2016ImplicationsVisualAttention,shimojo2003GazeBiasBoth}, where dwell times to AOIs belonging to the ultimately chosen alternative are longer across the trial (and increasingly so throughout the trial, indicated by the positive interaction term with time). Dwell times to decoys decreased significantly during the trial, and across the full trial, dwell times to decoys were shorter than other alternatives. Similarly, dwell times towards probability attributes p shortened across the trial. Across the full trial, however, dwell times towards probability AOIs were not shorter than those to outcome AOIs. Finally, dwell times to target alternatives were longer across the trial in both compromise and attraction trials. In addition, this effect increased throughout the trial in compromise, but not in attraction trials. Note that these effects are independent of the effect of choice, as choice is a separate predictor in the model. We could not find an association between the attribute rank (being the worst, best, or intermediate item on an attribute) and dwell time.

>>>>

While current gaze-dependent accumulation models of risky choice assume that each alternative can be assigned a singular underlying scale value (e.g., expected utilities for each alternative), choices are also assumed to be constructed in an accumulation process that remains malleable due to its dependence on gaze. Therefore, gaze-dependent accumulation models potentially could account for violations of IIA if the distribution of gaze itself changed depending on the decision context (e.g., shift towards dominant alternatives in the attraction effect, compromise alternatives in the attraction effect, or dissimilar alternatives in the similarity effect). Initial evidence for such an attention-shift in the attraction effect comes from a study of intertemporal choice \autocite{marini2020AttractionComesMany}. Furthermore, recent work has suggested gaze-dependent evidence accumulation as a model of multi-alternative, multi-attribute choice in the absence of context effects \autocite{cohen2017MultiattributeMultialternativeModels,gluth2018ValuebasedAttentionalCapture}.

@@

\subsubsection*{Attribute-wise gaze discount}
The analysis also included the option of attribute-wise gaze-dependent discounting (similar to the two-layer model from \citeauthor{glickman2019FormationPreferenceRisky} \autocite{glickman2019FormationPreferenceRisky} and the model presented in \citeauthor{fisher2017AttentionalDriftDiffusion} \autocite{fisher2017AttentionalDriftDiffusion}). If switched on, stimulus attributes of the currently unattended dimension (e.g., probability, when an lottery outcome was fixated) were discounted by a free parameter $\eta$ ($0 \le \eta \le 1$). In combination with \emph{additive} attribute integration, the attribute-wise gaze discount was applied after attribute normalization, but prior to the weighted addition. For \emph{multiplicative} attribute integration, attributes were discounted before entering probability-weighting and utility functions $w$ and $U$.

@@

Regression weight estimates are shown in Supplementary Fig. \ref{fig:dwell-regression-weights}. Across the trial, we find strong effects of position on dwell time, so that dwell times towards the top and left were longer. Significant negative interaction effects of the row and column predictors with time showed that these effects diminish across the trial. We also find a gaze-cascade effect \autocite{mullett2016ImplicationsVisualAttention,shimojo2003GazeBiasBoth}, where dwell times to AOIs belonging to the ultimately chosen alternative are longer across the trial (and increasingly so throughout the trial, indicated by the positive interaction term with time). Dwell times to decoys decreased significantly during the trial, and across the full trial, dwell times to decoys were shorter than other alternatives. Similarly, dwell times towards probability attributes $p$ shortened across the trial. Across the full trial, however, dwell times towards probability AOIs were not shorter than those to outcome AOIs. Finally, dwell times to target alternatives were longer across the trial in both compromise and attraction trials. In addition, this effect increased throughout the trial in compromise, but not in attraction trials. Note that these effects are independent of the effect of choice, as choice is a separate predictor in the model. We could not find an association between the attribute rank (being the worst, best, or intermediate item on an attribute) and dwell time.
